A Proposal for Linguistic Similarity Datasets Based on Commonality Lists

### 2016-06-14 paper.tex

% \title{Treating Similarity with Respect: How to Evaluate Models of Meaning?}
\title{A Proposal for Multi-category Lexical and Sentential Similarity Datasets Based on Commonality Lists}

@@

\begin{abstract}
Similarity is a core notion that is used in psychology, theoretical and computational linguistics. The similarity datasets that come from the two fields differ in design: psychological datasets are focused around a certain topic such as fruit names; linguistic datasets contain words from various categories. The later makes humans assign low similarity scores to the words that have nothing in common and to the words that have contrast in meaning, making similarity scores ambiguous. In this work we discuss the similarity collection procedure for a multi-category dataset that avoids score ambiguity and suggest changes to the evaluation procedure to reflect the insights of psychological literature for word, phrase and sentence similarity. We suggest to ask humans to provide a list of commonalities and differences instead of numerical similarity scores and employ the structure of human judgements beyond pairwise similarity. We believe that the proposed approach will give rise to datasets that test meaning representation models more thoroughly with respect to the human treatment of similarity.
\end{abstract}

@@

Similarity is the degree of resemblance between two objects or events \cite{WCS:WCS1282} and plays a crucial role in psychological theories of knowledge and behaviour, where it is used to explain such phenomena as classification and conceptualisation. \textit{Fruit} is a \emph{category} because it is a practical generalisation. Fruits are sweet and constitute deserts, so when one is presented with an unseen fruit, one can hypothesise that it is served toward the end of a dinner.

@@

% TODO: Mention \cite{bengio2006}?

@@

Both MEN and SimLex-999 provide pairs that do not share any similarity to control for false positives, and they do not control for the comparison scale. This makes similarity judgements ambiguous as it is not clear what low similarity values mean: incomparable notions or contrast in meaning. SimLex-999 assigns low similarity scores to the incomparable pairs (0.48, \textit{trick} and \textit{size}) and to antonymy (0.55, \textit{smart} and \textit{dumb}), but \textit{smart} and \textit{dumb} have relatively much more in common than \textit{trick} and \textit{size}!

@@

To build a similarity dataset which contains non-overlapping categories, one needs to avoid comparison of incomparable pairs. However, that itself requires an a priori knowledge of item similarity or belongingness to a category, making the problem circular. To get out of this vicious circle, one might erroneously  refer to an already existing taxonomy such as WordNet \cite{Miller:1995:WLD:219717.219748}. But in case of similarity, as \newcite{turney2012domain} points out, categories that emerge from similarity judgements are different from taxonomies. For example, \textit{traffic} and \textit{water} might be considered to be similar because of a functional similarity exploited in hydrodynamic models of traffic, but their lowest common ancestor in WordNet is \textit{entity}.

Since there is no way of deciding upfront whether there is a similarity relation between two words, the data collection procedure needs to test for both: relation existence and its strength. Numerical values, as has been shown in the introduction, do not fit this role, because of ambiguity. One way of avoiding the issue is to avoid asking humans for numerical similarity judgements, but instead to ask them to list commonalities and differences between the objects. As one might expect, similarity scores correlate with the number of listed commonalities \cite{markman1991commonalities,Markman1996,medin1993respects}. For incomparable pairs, the commonality list should be empty, but the differences will enumerate properties that belong to one entity, but not another \cite{markman1991commonalities,medin1993respects}. Verbally produced features (norms) for empirically derived conceptual representation of \newcite{McRae2005} is a good example of what data should be collected. But in contrast to \newcite{McRae2005}---where explicit comparison of concepts was avoided---participants should be asked to produce commonalities as part of similarity comparison.

@@

So far we have proposed a similarity judgement collection method that is robust to incomparable pairings. It also naturally gives rise to categories, because the absence of a relation between two entries means the absence of a common category. It still needs to be decided what words to include to the dataset.

@@

\paragraph{Prototype effect}

@@

\paragraph{Similarity in Context}

@@

\paragraph{Attention Attraction}

@@

\paragraph{Grammaticalisaton and Lexicalisaton}

@@

\paragraph{Discourse Fit}

@@

\paragraph{Language Sub-systems and Strategies}

>>>>

% \title{Treating Similarity with Respect: How to Evaluate Models of Meaning?}
\title{A Proposal for Linguistic Similarity Datasets Based on Commonality Lists}

@@

\begin{abstract}
Similarity is a core notion that is used in psychology and two branches of linguistics: theoretical and computational. The similarity datasets that come from the two fields differ in design: psychological datasets are focused around a certain topic such as fruit names, while linguistic datasets contain words from various categories. The later makes humans assign low similarity scores to the words that have nothing in common and to the words that have contrast in meaning, making similarity scores ambiguous. In this work we discuss the similarity collection procedure for a multi-category dataset that avoids score ambiguity and suggest changes to the evaluation procedure to reflect the insights of psychological literature for word, phrase and sentence similarity. We suggest to ask humans to provide a list of commonalities and differences instead of numerical similarity scores and employ the structure of human judgements beyond pairwise similarity for model evaluation. We believe that the proposed approach will give rise to datasets that test meaning representation models more thoroughly with respect to the human treatment of similarity.
\end{abstract}

@@

Similarity is the degree of resemblance between two objects or events \cite{WCS:WCS1282} and plays a crucial role in psychological theories of knowledge and behaviour, where it is used to explain such phenomena as classification and conceptualisation. \textit{Fruit} is a \emph{category} because it is a practical generalisation. Fruits are sweet and constitute deserts, so when one is presented with an unknown fruit, one can hypothesise that it is served toward the end of a dinner.

@@

@@

Both MEN and SimLex-999 provide pairs that do not share any similarity to control for false positives, and they do not control for the comparison scale. This makes similarity judgements ambiguous as it is not clear what low similarity values mean: incomparable notions, contrast in meaning or even the difference in comparison context. SimLex-999 assigns low similarity scores to the incomparable pairs (0.48, \textit{trick} and \textit{size}) and to antonymy (0.55, \textit{smart} and \textit{dumb}), but \textit{smart} and \textit{dumb} have relatively much more in common than \textit{trick} and \textit{size}!

@@

To build a similarity dataset that contains non-overlapping categories, one needs to avoid comparison of incomparable pairs. However, that itself requires an a priori knowledge of item similarity or belongingness to a category, making the problem circular. To get out of this vicious circle, one might erroneously  refer to an already existing taxonomy such as WordNet \cite{Miller:1995:WLD:219717.219748}. But in case of similarity, as \newcite{turney2012domain} points out, categories that emerge from similarity judgements are different from taxonomies. For example, \textit{traffic} and \textit{water} might be considered to be similar because of a functional similarity exploited in hydrodynamic models of traffic, but their lowest common ancestor in WordNet is \textit{entity}.

Since there is no way of deciding upfront whether there is a similarity relation between two words, the data collection procedure needs to test for both: relation existence and its strength. Numerical values, as has been shown in the introduction, do not fit this role, because of ambiguity. One way of avoiding the issue is to avoid asking humans for numerical similarity judgements, but instead to ask them to list commonalities and differences between the objects. As one might expect, similarity scores correlate with the number of listed commonalities \cite{markman1991commonalities,Markman1996,medin1993respects}. For incomparable pairs, the commonality list should be empty, but the differences will enumerate properties that belong to one entity, but not another \cite{markman1991commonalities,medin1993respects}. Verbally produced features (norms) for empirically derived conceptual representation of \newcite{McRae2005} is a good example of what and how the data should be collected. But in contrast to \newcite{McRae2005}---where explicit comparison of concepts was avoided---participants should be asked to produce commonalities as part of similarity comparison.

@@

So far, we have proposed a similarity judgement collection method that is robust to incomparable pairings. It also naturally gives rise to categories, because the absence of a relation between two entries means the absence of a common category. It still needs to be decided what words to include to the dataset.

@@

\paragraph{Prototype effects}

@@

\paragraph{Similarity in context}

@@

\paragraph{Attention attraction}

@@

\paragraph{Grammaticalisaton and lexicalisaton}

@@

\paragraph{Discourse fit}

@@

\paragraph{Language Sub-systems and strategies}
